Supermartingales and Termination Analysis of Statistical PCF

### 2021-01-09 LICS21/confluent-proofs.tex
Trim down non-appendix size.

% !TEX root = main.tex
The rules defining $\sim_c$ can be more intuitively understood as diagrams. In the following diagrams, a circle represents a term (or skeleton), and the leaves on the trees are subterms at the positions indicated by the labels in the other nodes. The labels on the arrows between trees are the positions of the reductions, and dashed arrows represent some number of reductions in a row. If a reduction sequence includes one branch of one of these cases at some point in it, then the positions in the reduction sequence obtained by substituting in the other branch are all related by $\sim_c$ to the corresponding positions in the original reduction sequence.

\lo{TODO. Remove the white space by repositioning the sim diagrams.}
\include{sim-diagram}
%The diagram is not actually in the figure environment because that doesn't work properly with it extending over several pages. The empty figure is just needed to get the label and caption.
\begin{figure}[htb]
\caption{Illustration of the $\sim_c$ rules. \label{sim-diagram}
}
\end{figure}

@@

\AstEquivalence*
\begin{proof}
\akr{I'm not sure whether this proof is sufficiently detailed. It would be pretty easy to fill it in more, just very verbose.}

For both $\red$ and $\Rightarrow_{\cbv}$, the probability measure on the sample space can be used to define a measure on the reduction sequences of each finite length. In the case that the reduction sequences terminate after a finite number of steps by reaching a value, the reduction sequence defined using $\Rightarrow_{\cbv}$ should be taken to continue by repeating the last term indefinitely, as $\red$ does, even though $\Rightarrow_{\cbv}$ specifies no next step in this case. The distributions are equal for each length, by induction, as follows.

@@

For the inductive case, suppose that the distributions of reduction sequences of length $n$ are equal. \akr{TODO: mention the case that it has already terminated by step $n$, or is about to. The two versions of the semantics handle it a little differently.}

@@

>>>>

% !TEX root = main.tex
\label{apx:confluent}
The rules defining $\sim_c$ can be more intuitively understood as diagrams. In the following diagrams, a circle represents a term (or skeleton), and the leaves on the trees are subterms at the positions indicated by the labels in the other nodes. The labels on the arrows between trees are the positions of the reductions, and dashed arrows represent some number of reductions in a row. If a reduction sequence includes one branch of one of these cases at some point in it, then the positions in the reduction sequence obtained by substituting in the other branch are all related by $\sim_c$ to the corresponding positions in the original reduction sequence.
\include{sim-diagram}
%The diagram is not actually in the figure environment because that doesn't work properly with it extending over several pages. The empty figure is just needed to get the label and caption.
\begin{figure}[htb]
\caption{Illustration of the $\sim_c$ rules. \label{sim-diagram}
}
\end{figure}

@@

\AstEquivalence*
\begin{proof}
For both $\red$ and $\Rightarrow_{\cbv}$, the probability measure on the sample space can be used to define a measure on the reduction sequences of each finite length. In the case that the reduction sequences terminate after a finite number of steps by reaching a value, the reduction sequence defined using $\Rightarrow_{\cbv}$ should be taken to continue by repeating the last term indefinitely, as $\red$ does, even though $\Rightarrow_{\cbv}$ specifies no next step in this case. The distributions are equal for each length, by induction, as follows.

@@

For the inductive case, suppose that the distributions of reduction sequences of length $n$ are equal.

@@

\Cref{ex:sum of powers}
Recall that
\begin{align*}
M =&\ (\lambda n. P[n]) \lfloor \tsample^{-1/2} \rfloor \\
P[n] =&\ (\lambda p. \Xi[p] n) (\lambda x. \Theta[x] n)) \\
\Xi[p] =&\ \tY \lambda f n. \tif{n = 0}{\underline 0}{p n + f (n-1)} \\
\Theta[x] =&\ \tY \lambda f n. \tif{n = 0}{\underline 1}{x \times f (n-1)}.
\end{align*}

Let $r$ be the reduction strategy which reduces $M$ in the order
\begin{align*}
&\ M \\
\to^* &\ P[\underline n] \\
\to^* &\ (\lambda p.\ \Xi[p]\ \underline n)\ (\lambda x. \underbrace{x \times \dots \times x \times}_n \underline 1) \\
\to^{\phantom *} &\ \Xi[\lambda x. \underbrace{x \times \dots \times x \times}_n \underline 1]\ \underline n \\
\to^{\phantom *} &\ \tif{\underline n = 0}{\underline 0}{\underline{n^n} + \Xi[\lambda x. x \times \dots \underline 1]\ \underline{n-1}} \\
\to^* &\ \underline{n^n} + \underline{(n-1)^n} + \dots + \underline 0 \\
\to^* &\ \underline{\sum_{k=0}^n k^n}
\end{align*}
(this is a little underspecified, but the details are not important). A suitable sparse ranking function for it is
\begin{align*}
M \mapsto & \frac{\pi^2}{3} \\
P[\underline n] \mapsto & 2n+2 \\
(\lambda p.\ \Xi[p]\ \underline n)\ (\lambda x. \underbrace{x \times \dots \times x \times}_{n-k} (\Theta[x]\ \underline k)) \mapsto & n + k + 2 \\
\Xi[\lambda x. \underbrace{x \times \dots \times x \times}_n \underline 1]\ \underline n \mapsto & n + 1 \\
\underline{n^n} + \dots + \underline{(n-k+1)^n} + \Xi[\lambda x. x \times \dots \underline 1]\ \underline k \mapsto & k + 1 \\
\underline{\sum_{k=0}^n k^n} \mapsto & 0.
\end{align*}

The first step is justified because the probability of reaching $P[\underline n]$ for any specific $n \geq 0$ is $(n+1)^{-2} - (n+2)^{-2} = \frac{2n + 3}{(n+1)^2(n+2)^2}$, therefore the expected next value of the ranking function is $\sum_{n=0}^{\infty} \frac{(2n+2)(2n + 3)}{(n+1)^2(n+2)^2} = \frac{\pi^2}{3}$. The other steps are deterministic, and involve 1 or 0 $\tY$-reductions each. Given that all of the reduction steps after the first are deterministic, and the number of $\tY$ reduction steps is not that hard to count, defining the sparse ranking function's values only at $M$ and $P[\underline n]$ would also have been reasonable, although in a similar term which had more random samples throughout, that would not have been so simple.

Providing a ranking function of some sort for this term in a similar level of detail would have been rather more complicated using only the standard reduction strategy. The recursion in $\Theta$ would have had to be evaluated separately for every time it was used, so there would have been more (and more complex) terms to consider in the reduction sequence. Also, the number of $\tY$-reductions from $P[\underline n]$ would have been $n^2 + 2n + 1$ instead of $2n + 2$, therefore the expected number of $\tY$ reductions starting from $M$ would be infinite, i.e.~$M$ is not $\tY$-PAST, therefore it is not even rankable. It would be possible to define an antitone sparse ranking function for it, but it would have to include terms such as
\begin{align*}
& \underline{n^n} + \dots + \underline{(n-k_1+1)^n} + \\
& (\underbrace{\underline{n-k_1} \times \dots \times \underline{n-k_1} \times}_{n-k_2-1} (\Theta[\underline{n-k_1}] (\underline{k_2} - 1))) + \\
& \Xi[\lambda x. \Theta[x] \underline n] (\underline{n-k_1} - 1) \mapsto \log((n+1)k_1 + k_2)
\end{align*}
as only one of the several cases, which is significantly more complicated and difficult to use.

It wouldn't even be possible in this case to give a sparser version of the ranking function defined only at $M$ and $P[\underline n]$, because the amount that an antitone sparse ranking function must decrease at each $\tY$-reduction step depends on the value of the ranking function at the next term where it is defined, therefore it is necessary to have these intermediate steps in order for the ranking function to be able to change sufficiently slowly.

### 2021-01-09 LICS21/ranking-proofs.tex
Trim down non-appendix size.

% !TEX root = main.tex
\section{Constructing ranking functions}
\label{sec:ranking}

Although rankability implies almost sure termination, the converse does not hold in general. For example,
\begin{equation}
\tif{{-}\tsample < 0}{\underline{0}}{(\tY \lambda x. x) \, \underline 0}
\label{ex:0 probability reachable}
\end{equation}
terminates in 3 steps with probability 1, but isn't rankable because $(\tY \lambda x. x) \, \underline 0$ is reachable, although that has probability 0. 
Not only is this counterexample AST, it's PAST. %positively almost surely terminating i.e.~the expected time to termination is finite.

A ranking function can be constructed under the stronger assumptions that, for every $N$ reachable from $M$, the expected number of $\tY$-reduction steps from $N$ to a value is finite. 
In particular, the expected number of $\tY$-reduction steps from each reachable term is a ranking function. 
However, note that a finite number of expected $\tY$-reduction steps does not necessarily imply a finite number of expected total reduction steps.

\begin{example}
\label{ex:tY finite does not imply t finite}
The term $M = \Xi \, (\lambda x. x+1)$ where
\[
\Xi := \tY \lambda f \, n. \tif{\tsample - 0.5 < 0}{n \, \underline 0}{f (\lambda x. n (n \, x))}
\] 
terminates with only 2 $\tY$-reductions on average, i.e.,~$\expect{T_M^\tY} = 2$, but applies the increment function $2^n$ times with probability $2^{-n-1}$ for $n \geq 0$, which diverges, i.e.,~$\expect{T_M} = \infty$.
\end{example}

\begin{theorem} \label{thm:minimal}
Given a closed term $M$, the function $f:Rch(M) \to \mathbb R$ given by 
\[
f(N) := \mathbb E [\text{number of }\tY\text{-reduction steps from }N\text{ to a value}]
\] 
if it exists, is the least of all possible ranking functions of $M$.
\end{theorem}
\begin{proof}
Let $f$ be the candidate least ranking function defined above, and suppose $g$ is another ranking function such that $f(N) > g(N)$ for some $N \in Rch(M)$. The restrictions of $f$ and $g$ to $Rch(N)$ have the same properties assumed of $f$ and $g$, so assume w.l.o.g.~that $N=M$. The difference $g - f$ is then a supermartingale (with the same setup as in the proof of %\Cref{thm:rankable gives supermartingale}
%Theorem \ref{rankable implies ast}) 
\Cref{thm:rankable and strict rankable});
therefore $\mathbb E[g(M_n)] \leq \mathbb E[f(M_n)] + g(M)-f(M)$, for all $n$.
Now $\mathbb E[f(M_n)] = \sum_{k=n}^\infty \mathbb P[M_k = E[\tY N] \text{ for some }E, N] \to 0 \text{ as } n \to \infty$; 
therefore as $g(M) - f(M) < 0$, eventually $\mathbb E[g(M_n)] < 0$, which is impossible. 
It follows that $g \geq f$ as required.

@@

\subsection{Sparse ranking functions}
Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and \Cref{thm:minimal} makes even stronger assumptions than almost sure termination, so it isn't useful for proving it.

\begin{example}[Geometric distribution]
% Consider the term
% \begin{align*}
% &(\tY \lambda f, n. \\
% &\quad \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
% &) \, \underline{0}
% \end{align*}
%\[
%\big(\tY \lambda f, n. \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
%\big) \, \underline{0}
%\]
Let
\[
\Theta := \lambda f, n. \tif{\tsample - 0.5 < 0}{n}{f (n+1)}. 
\]
The term  $(\tY \, \Theta) \, \underline 0$ generates a geometric distribution.
\iffalse
\lo{\Cref{fig:geometric distribution} is confusing: the arrow means either one-step reduction, or its reflexive transitive closure, or ``has the same syntactic shape''.}

\akr{The arrows all mean the one-step reduction $\to$, I just didn't intend for the values of $i$ and $r$ to be consistent between the terms. I would much rather have the diagram be complete (with all the reductions and all the terms).}

\lo{Perhaps you have in mind an informal, schematic interpretation, reading a term $C[i, r]$ as ``there exists a term $C[i, r]$, for appropriate $i, r \in \mathbb{R}$''.
One way to make this precise is to consider the graph obtained by quotienting the one-step reduction relation by an appropriate equivalence relation over terms (that specifies this scheme reading).}

\lo{\Cref{fig:geometric distribution 1} is less ambitious, but it is not confusing in that arrow does mean one-step reduction (also note $r'$), and it would be correct to say (cf.~the preceding blue-highlighted phrase): ``Despite its simplicity, $\mathit{Rch}(\tY \, \Theta \, \underline 0)$ contains all the terms in \Cref{fig:geometric distribution 1}, for each $i \geq 0$''.}

\lo{The graph can be simplified somewhat if we redefine general recursion using a fixpoint operator $\tY_{f, n}\Theta$ with redex rule as per \Cref{eq:letrec redex}: see \Cref{fig:geometric distribution 2}.
But (especially in view of the vast investment in the confluent trace semantics) it is perhaps too late to change things now?}
\fi
Despite its simplicity, its $\mathit{Rch}$ contains all the terms in \Cref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$.
%\lo{TODO: reduce the figure to avoid overflow.}\akr{It still overflows the gutter, but not the actual text of the other column.}\lo{OK. Thanks.} 
The meanings of the substitutions (to be applied to the contractum) and conditions that label some of the reductions in the diagram should be self-explanatory.

@@

\iffalse
\begin{figure}
\centering
\begin{tikzpicture}[node distance=0.5em, anchor=west]
    \node (1) at (0,-0) {$\tY\, \Theta\, \underline i$ \quad [provided $i = 0$]};
    \node (2) at (0,-1) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, \underline i$};
    \node (3) at (0,-2) {$(\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})})\, \underline i$};
    \node (4) at (0,-3) {$\tif{\tsample \underline{- 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
    \node (5) at (0,-4) {$\tif{\underline r\, \underline{- 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
    \node (6) at (0,-5) {$\tif{\underline r' < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
    \node (7) at (2,-6) [text width=] {$\underline i$};
    \node (8) at (0,-7) {$\tY\, \Theta\, (\underline i\, \underline{+ 1})$};
    \node (9) at (0,-8) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, (\underline i\, \underline{+ 1})$};
    \node (10) at (0,-9) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, \underline{i + 1}$};

    \draw [->] (1.south west) to [bend right=45] (2.north west);
    \draw [->] (2.south west) to [bend right=45] (3.north west);
    \draw [->] (3.south west) to [bend right=45] (4.north west);
    \draw [->] (4.south west) to [bend right=45] (5.north west);
    \draw [->] (5.south west) to [bend right=45] (6.north west);
    \draw [->] (6.south west) to [bend right=45] (8.north west);
    \draw [->] (8.south west) to [bend right=45] (9.north west);
    \draw [->] (9.south west) to [bend right=45] (10.north west);
    \draw [->] (6) to (7);
\end{tikzpicture}
\caption{Some reachable terms of $\tY \, \Theta \, \underline 0$\label{fig:geometric distribution 1}}
\end{figure}

\begin{figure}
\centering
\begin{tikzpicture}[node distance=0.5em, anchor=west]
    \node (1) at (0,-0) {$\tY_{f, n}\Theta\, \underline i$};
    %\node (2) at (0,-1) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY_{f, n}\Theta\, (n \underline{+ 1})}) z)\, \underline i$};
    \node (3) at (0,-2) {$(\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY_{f, n}\Theta\, (n \underline{+ 1})})\, \underline i$};
    \node (4) at (0,-3) {$\tif{\tsample \underline{- 0.5} < 0}{\underline i}{\tY_{f, n}\Theta\, (\underline i\, \underline{+ 1})}$};
    \node (5) at (0,-4) {$\tif{\underline r\, \underline{- 0.5} < 0}{\underline i}{\tY_{f, n}\Theta\, (\underline i\, \underline{+ 1})}$};
    \node (6) at (0,-5) {$\tif{\underline r' < 0}{\underline i}{\tY_{f, n}\Theta\, (\underline i\, \underline{+ 1})}$};
    \node (7) at (2,-6) [text width=] {$\underline i$};
    \node (8) at (0,-7) {$\tY_{f, n}\Theta\, (\underline i\, \underline{+ 1})$};
    \node (9) at (0,-8) {$\tY_{f, n}\Theta\, \underline{i + 1}$};
    %\node (9) at (0,-8) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY_{f, n}\Theta\, (n \underline{+ 1})}) z)\, (\underline i\, \underline{+ 1})$};
    %\node (10) at (0,-9) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY_{f, n}\Theta\, (n \underline{+ 1})}) z)\, \underline{i + 1}$};

    \draw [->] (1.south west) to [bend right=45] (3.north west);
    %\draw [->] (2.south west) to [bend right=45] (3.north west);
    \draw [->] (3.south west) to [bend right=45] (4.north west);
    \draw [->] (4.south west) to [bend right=45] (5.north west);
    \draw [->] (5.south west) to [bend right=45] (6.north west);
    \draw [->] (6.south west) to [bend right=45] (8.north west);
    \draw [->] (8.south west) to [bend right=45] (9.north west);
    %\draw [->] (9.south west) to [bend right=45] (10.north west);
    \draw [->] (6) to (7);
\end{tikzpicture}
\caption{Some reachable terms of $\tY_{f, n}\Theta \, \underline 0$\label{fig:geometric distribution 2}}
\end{figure}
\fi
\end{example}

Even in this simple case, defining a ranking function explicitly is awkward because of the number of cases, although in most cases, because the value need only be greater than or equal to that of the next term in sequence, it suffices to take the ranking function as having the same value as the next term, so that overall it takes only 3 distinct values.
(We will explain why later.)

The definition of rankability is also inconvenient for syntactic sugar. It could be useful, for example, to define $M \oplus_p N := \tif{\tsample - p < 0} M N$, where $M \oplus_p N$ reduces to $M$ or $N$, depending on the first value of $s \in \entrosp$, with probability $p$ resp.~$(1-p)$. Technically though, it reduces first to $\tif{\underline r - p < 0} M N$ for all $r \in I$, so those terms all need values of the ranking function too.

%\paragraph{}
In both of these cases, there are only some values of the ranking function that are semantically important. 
\begin{definition}
Define a \emph{sparse ranking function} on a closed term $M$ to be a partial function $f : Rch(M) \rightharpoonup \mathbb R$ such that
\begin{itemize}
    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
    \item $f$ is defined at $M$.
    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually \akr{It would be nicer for this to be only almost certain, to make this more neatly in-between rankability and AST, but that doesn't actually work because of terms reachable with 0 probability. Of course, ASTness doesn't imply rankability, so any definition of sparse ranking function that includes all Y-PAST terms wouldn't always extend to a total ranking function.} reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
\end{itemize}
\end{definition}
A sparse ranking function that is total is just a ranking function. Providing a sparse ranking function is essentially part way between providing a ranking function and directly proving almost sure termination.

\begin{theorem} \label{thm:partial implies rankable}
Every sparse ranking function is a restriction of a ranking function.
\end{theorem}
\begin{proof}
Take a closed term $M$ and sparse ranking function $f$ on $M$. Define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
\begin{align*}
    f_1(N) & := f(N) \text{ whenever $f(N)$ is defined,} \\
    f_1(V) & := 0 \text{ for values $V$ not in the domain of $f$.}
\end{align*}

Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) := \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$. 
The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on sparse ranking functions. Define $f_2(N) = \int_\entrosp f_1(\nnext(N,s)) + g(N,s) \, \mu_{\entrosp}(\mathrm d s)$. The (total) function $f_2$ agrees with $f$ on $f$'s domain, and it is a ranking function on $M$ (in fact, the least ranking function of which $f$ is a restriction).
\lo{TODO. The preceding sentence deserves a proof.}
\end{proof}

As a corollary, any term which admits a sparse ranking function terminates almost surely.

\subsection{Examples}
Let $M \oplus_p N = \tif{\tsample - p < 0} M N$, for $p \in (0,1]$, then there are the pseudo-reduction relations
\begin{align*}
E[M \oplus_p N] \to^3 & E[M] & \\
E[M \oplus_p N] \to^3 & E[N] & \\
\red^3(E[M \oplus_p N], s) = & \left\{
    \begin{array}{ll}
        (E[M],\pi_t(s)) & \text{if } \pi_h(s) < p \\
        (E[N],\pi_t(s)) & \text{if } \pi_h(s) \geq p. \\
    \end{array} \right .
\end{align*}
A sparse ranking function could be defined with respect to this shortcut reduction simply by replacing $\to$ and $\red$ in the definition of a sparse ranking function by a version that goes straight from $N \oplus_p O$ to $N$ or $O$. Such a pseudo-sparse ranking function would then be a partial function from a subset of $Rch(M)$, so it could also be considered as a partial function from all of $Rch(M)$, and it would in fact also be an actual sparse ranking function. It is therefore possible to prove rankability directly using the shortcutted reductions.

A similar procedure would work for other forms of syntactic sugar. If a closed term $N$ eventually reduces to one of a set of other terms $\{N_i \mid i \in I\}$ with certain probabilities, a sparse ranking function defined with respect to a reduction sequence that skips straight from $N$ to $N_i$ is also a valid sparse ranking function for the original reduction function, and therefore its existence implies almost sure termination. There is a caveat, however, that $\tY$-reduction steps skipped over in the shortcut still need to be counted for the expected number of $\tY$-reduction steps.

%\paragraph{}
With this abbreviation, the geometric distribution example from earlier can be written as $(\tY \lambda f n.\ n \oplus_{0.5} f(n+1)) \, \underline 0$. It is then easy to see that the following is a sparse ranking function:
\begin{align*}
(\tY \lambda f n.\ n \oplus_{0.5} f(n+1))\, N & \mapsto 2 \\
\underline i \oplus_{0.5} (\tY \lambda f n.\ n \oplus_{0.5} f(n+1))\, (\underline i + 1) & \mapsto 1 \\
\underline i & \mapsto 0.
\end{align*}

In fact, even the partial function $\tY \Theta N \mapsto 2$ alone is a sparse ranking function for this term.

>>>>

% !TEX root = main.tex
\ThmMinimal*
\begin{proof}
Let $f$ be the candidate least ranking function defined above, and suppose $g$ is another ranking function such that $f(N) > g(N)$ for some $N \in Rch(M)$. The restrictions of $f$ and $g$ to $Rch(N)$ have the same properties assumed of $f$ and $g$, so assume w.l.o.g.~that $N=M$. The difference $g - f$ is then a supermartingale (with the same setup as in the proof of %\Cref{thm:rankable gives supermartingale}
%Theorem \ref{rankable implies ast}) 
\Cref{thm:rankable and strict rankable});
therefore $\mathbb E[g(M_n)] \leq \mathbb E[f(M_n)] + g(M)-f(M)$, for all $n$.
Now $\mathbb E[f(M_n)] = \sum_{k=n}^\infty \mathbb P[M_k = E[\tY N] \text{ for some }E, N] \to 0 \text{ as } n \to \infty$; 
therefore as $g(M) - f(M) < 0$, eventually $\mathbb E[g(M_n)] < 0$, which is impossible. 
It follows that $g \geq f$ as required.

@@

@@

\PartialImpliesRankable*
\begin{proof}
Take a closed term $M$ and sparse ranking function $f$ on $M$. Define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
\begin{align*}
    f_1(N) & := f(N) \text{ whenever $f(N)$ is defined,} \\
    f_1(V) & := 0 \text{ for values $V$ not in the domain of $f$.}
\end{align*}

Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) := \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$. 
The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on sparse ranking functions. Define $f_2(N) = \int_\entrosp f_1(\nnext(N,s)) + g(N,s) \, \mu_{\entrosp}(\mathrm d s)$. The (total) function $f_2$ agrees with $f$ on $f$'s domain, and it is a ranking function on $M$ (in fact, the least ranking function of which $f$ is a restriction, by the same argument as \Cref{thm:minimal}).
\end{proof}
